The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-04-26 paper/ms.tex
Fix equation refs throughout

While the formula for $J_{\phi}$ is phase independent, $J_R$ and $J_z$ are
both phase dependent via Eq. \eqref{eq:orbits_epi}. We consider two cases. First, at the point of maximum
amplitude or when $(v_R, v_z) = (0, 0)$ and second, at the point of maximum
velocity $(R, z) = (R_g, 0)$. For $J_R$ we have that,
\beq\label{eq:JR_epi_final}
    \begin{aligned}
        J_R &= \kappa A_R^2 && \text{if $v_R = 0$} \\
    J_R &= \frac{v_{R,\text{max}}^2}{2\kappa^2} && \text{if $R=R_g$} \text{.}
    \end{aligned}
\eeq
For $J_z$, we have that,
\beq\label{eq:Jz_epi_final}
    \begin{aligned}
        J_z &= \kappa A_z^2 && \text{if $v_z = 0$} \\
        J_z &= \frac{v_{z,\text{max}}^2}{2\kappa^2} && \text{if $z=0$} \text{.}
    \end{aligned}
\eeq

@@

In each panel of Figure~\ref{fig:many_orbit_wrong_ref}, where relevant, we
include the estimation of the action errors derived under the epicyclic
approximation from Equation~\ref{eq:Ji_err_mosttime}. This Equation is relevant since during most of the orbit the star will be at maximum radial and vertical amplitude and we do not consider velocity offsets. Note that we consider an error
in the $x$-coordinate $\Delta x$, which is not quite the same as $\Delta R$.
For observations of stars close to us, we have that $\Delta x \sim \Delta R$,
but for the experiment performed in this Section we consider observations of
the star throughout its entire orbit. This introduces a factor of $2/\pi$ when
converting from $\Delta R$ to $\Delta x$, which we derive in
Appendix~\ref{app:deltax}.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_many_orbits.pdf}
\end{center}
\caption{We report one half the 95\uth{} minus 5\uth{} percentile of the error
in the measured action ($\Delta J_i$) from coordinate system errors for the
thin, thick, and halo orbits (Table~\ref{tab:orbits}). See discussion in the
text for the justification in using this to measure the magnitude of the
induced error. The {\em left}, {\em center}, and {\em right} panels show the
result for $J_R$, $J_{\phi}$, and $J_z$, respectively. The {\em upper} panels
consider an offset in $z$ and the {\em lower} panels consider an offset in $x$
(equivalently, an offset in the Solar radius). In some panels we also plot as
dashed lines the epicyclic prediction of the induced action error
(Equations~\ref{eq:actions_epi_error}~and~\ref{eq:Jr_epi_dx}). In the
epicyclic approximation, a $z$ offset only induces an error in $J_z$ --- for
all three orbits the epicyclic approximation is a good description of the
$J_z$ error. An $x$ offset induces an error in $J_R$ and $J_{\phi}$. The error in
$J_R$ is somewhat well-described for the thin-disk and thick-disk orbits, and
a poor description for the halo orbit. For $J_{\phi}$, the epicyclic approximation
is not a good description for any orbit.}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

@@

To further understand the effect of the midplane error, we also plot the
fractional error in $J_z$ as a function of $J_z$ for $z$ offsets of $10, 50,
100\,\pc$ (orange, cyan, and green, respectively) in
Figure~\ref{fig:dJz_fun_Jz}. For a thin disk-like orbit
($J_z\sim0.5\,\actunit$), even a $10\,\pc$ offset in $z$ is enough to induce a
$\sim25\%$ error in $J_z$. For larger values of $J_z$, the fractional errors
are suppressed, but the induced error can still be large depending on how
great the $z$ offset is. We also plot the epicylic prediction for $\Delta J_z
/ J_z$ from Equation~\ref{eq:Ji_err_mosttime} as dashed lines for each $z$
offset. We find that the epicyclic approximation matches the numerical
estimate quite well.

@@

The second issue, which is related to the point of this paper, is that based
on the results in our previous sections computing $\bm{J}_i$ for a field star
and directly comparing it to  $\bm{J}_c$ will have significant systematic
error if the cluster and field star are far enough apart to have significantly
different local midplanes (see Figures \ref{fig:midplane} and \ref{fig:midplane_chervin}). Since our criterion for cluster membership (distance in action space) scales with the
cluster mass, the region where the search will succeed varies with cluster
mass as well. To illustrate this effect, we assume a cluster mass $m_c$ and
estimate the expected action-space dispersion from
Equation~\ref{eq:action_disp}. We then consult
Figure~\ref{fig:many_orbit_wrong_ref} to determine the midplane offset
necessary in order for the induced error in the vertical action to be the same
as the expected action-space dispersion. We perform this analysis for the
thin-disk-like and thick-disk-like orbits considered in
Section~\ref{ssec:quant}, assuming that the halo-like orbit is unlikely for
members of known open clusters (see Table~\ref{tab:real_clusters}). The result
for the allowable midplane offset for these two orbits in shown in
Figure~\ref{fig:cluster_offset} as a function of cluster mass.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/cluster_offset.pdf}
\end{center}
\caption{The $z$ offset necessary in order for the action error induced by our
midplane effect to be comparable to the intrinsic action dispersion from a
cluster of mass $m_c$. We compute the intrinsic dispersion from
Equation~\ref{eq:action_disp} and convert the dispersion to a $z$ offset using
the result from Figure~\ref{fig:many_orbit_wrong_ref}. We consider the result
for the thin and thick orbit (Table~\ref{tab:orbits}). Nearby open clusters
(Table~\ref{tab:real_clusters}) have orbits closest to the thin orbit.}
\label{fig:cluster_offset}
\end{figure}

>>>>

While the formula for $J_{\phi}$ is phase independent, $J_R$ and $J_z$ are
both phase dependent via Equation~\eqref{eq:orbits_epi}. We consider two cases. First, at the point of maximum
amplitude or when $(v_R, v_z) = (0, 0)$ and second, at the point of maximum
velocity $(R, z) = (R_g, 0)$. For $J_R$ we have that,
\beq\label{eq:JR_epi_final}
    \begin{aligned}
        J_R &= \kappa A_R^2 && \text{if $v_R = 0$} \\
    J_R &= \frac{v_{R,\text{max}}^2}{2\kappa^2} && \text{if $R=R_g$} \text{.}
    \end{aligned}
\eeq
For $J_z$, we have that,
\beq\label{eq:Jz_epi_final}
    \begin{aligned}
        J_z &= \kappa A_z^2 && \text{if $v_z = 0$} \\
        J_z &= \frac{v_{z,\text{max}}^2}{2\kappa^2} && \text{if $z=0$} \text{.}
    \end{aligned}
\eeq

@@

In each panel of Figure~\ref{fig:many_orbit_wrong_ref}, where relevant, we
include the estimation of the action errors derived under the epicyclic
approximation from Equation~\eqref{eq:Ji_err_mosttime}. This Equation is relevant since during most of the orbit the star will be at maximum radial and vertical amplitude and we do not consider velocity offsets. Note that we consider an error
in the $x$-coordinate $\Delta x$, which is not quite the same as $\Delta R$.
For observations of stars close to us, we have that $\Delta x \sim \Delta R$,
but for the experiment performed in this Section we consider observations of
the star throughout its entire orbit. This introduces a factor of $2/\pi$ when
converting from $\Delta R$ to $\Delta x$, which we derive in
Appendix~\ref{app:deltax}.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_many_orbits.pdf}
\end{center}
\caption{We report one half the 95\uth{} minus 5\uth{} percentile of the error
in the measured action ($\Delta J_i$) from coordinate system errors for the
thin, thick, and halo orbits (Table~\ref{tab:orbits}). See discussion in the
text for the justification in using this to measure the magnitude of the
induced error. The {\em left}, {\em center}, and {\em right} panels show the
result for $J_R$, $J_{\phi}$, and $J_z$, respectively. The {\em upper} panels
consider an offset in $z$ and the {\em lower} panels consider an offset in $x$
(equivalently, an offset in the Solar radius). In some panels we also plot as
dashed lines the epicyclic prediction of the induced action error
(Equations~\eqref{eq:actions_epi_error}~and~\eqref{eq:Jr_epi_dx}). In the
epicyclic approximation, a $z$ offset only induces an error in $J_z$ --- for
all three orbits the epicyclic approximation is a good description of the
$J_z$ error. An $x$ offset induces an error in $J_R$ and $J_{\phi}$. The error in
$J_R$ is somewhat well-described for the thin-disk and thick-disk orbits, and
a poor description for the halo orbit. For $J_{\phi}$, the epicyclic approximation
is not a good description for any orbit.}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

@@

To further understand the effect of the midplane error, we also plot the
fractional error in $J_z$ as a function of $J_z$ for $z$ offsets of $10, 50,
100\,\pc$ (orange, cyan, and green, respectively) in
Figure~\ref{fig:dJz_fun_Jz}. For a thin disk-like orbit
($J_z\sim0.5\,\actunit$), even a $10\,\pc$ offset in $z$ is enough to induce a
$\sim25\%$ error in $J_z$. For larger values of $J_z$, the fractional errors
are suppressed, but the induced error can still be large depending on how
great the $z$ offset is. We also plot the epicylic prediction for $\Delta J_z
/ J_z$ from Equation~\eqref{eq:Ji_err_mosttime} as dashed lines for each $z$
offset. We find that the epicyclic approximation matches the numerical
estimate quite well.

@@

The second issue, which is related to the point of this paper, is that based
on the results in our previous sections computing $\bm{J}_i$ for a field star
and directly comparing it to  $\bm{J}_c$ will have significant systematic
error if the cluster and field star are far enough apart to have significantly
different local midplanes (see Figures \ref{fig:midplane} and \ref{fig:midplane_chervin}). Since our criterion for cluster membership (distance in action space) scales with the
cluster mass, the region where the search will succeed varies with cluster
mass as well. To illustrate this effect, we assume a cluster mass $m_c$ and
estimate the expected action-space dispersion from
Equation~\eqref{eq:action_disp}. We then consult
Figure~\ref{fig:many_orbit_wrong_ref} to determine the midplane offset
necessary in order for the induced error in the vertical action to be the same
as the expected action-space dispersion. We perform this analysis for the
thin-disk-like and thick-disk-like orbits considered in
Section~\ref{ssec:quant}, assuming that the halo-like orbit is unlikely for
members of known open clusters (see Table~\ref{tab:real_clusters}). The result
for the allowable midplane offset for these two orbits in shown in
Figure~\ref{fig:cluster_offset} as a function of cluster mass.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/cluster_offset.pdf}
\end{center}
\caption{The $z$ offset necessary in order for the action error induced by our
midplane effect to be comparable to the intrinsic action dispersion from a
cluster of mass $m_c$. We compute the intrinsic dispersion from
Equation~\eqref{eq:action_disp} and convert the dispersion to a $z$ offset using
the result from Figure~\ref{fig:many_orbit_wrong_ref}. We consider the result
for the thin and thick orbit (Table~\ref{tab:orbits}). Nearby open clusters
(Table~\ref{tab:real_clusters}) have orbits closest to the thin orbit.}
\label{fig:cluster_offset}
\end{figure}
